Efficient ancestry and mutation simulation with msprime 1.0

### 2021-11-24 response-to-reviewers.tex

\begin{reply}
We have rephrased this to say "(see also Kelleher et al., 2016, Fig. 2)".
\end{reply}

\begin{point}
Line 370 - population size(s)
\end{point}

\begin{reply}
Thank you, we have corrected this sentence in the revised manuscript.
\end{reply}

\begin{point}
Line 420 - Despite the long description, I was not able to follow how the
previous paragraph implies that "work is spread out relatively evenly on the
coalescent time scale"
\end{point}

\begin{reply}
This is implied by the fact that the number of lineages decreases slowly
on the coalescent timescale. However, the point is not crucial to the
main goal of the paragraph, and we have therefore deleted the sentence
for clarity.
\end{reply}

\begin{point}
Line 422 - There is no context to what is meant by "large" or "small" here - is
2 or 10 or 10000 "large".
\end{point}

### 2021-11-24 response-to-reviewers.tex

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication. We are glad that
Reviewers 1 and 2 are enthusiastic about this community effort, and their reviews
have helped to resolve a number of issues with the manuscript. We are
disappointed that although Reviewer 3 agrees in the value of the software,
they do not believe the manuscript
is appropriate for publication in Genetics. However, as we point out in the
response, it is unclear whether they were reviewing according to the
``Methods, technology, and resources`` section scope.

\subsection*{Specific comments}
% We address your specific comments below:
% % \reviewersection

@@

\textit{
It is most important that you address the following in your resubmission:
Improve the structure and focus of the main body of the manuscript. This should
include moving a significant portion of the material to appendices or
supplemental material, and adding an up-front summary of the main new features
of msprime (relative to the 2016 paper).}

We have shortened the manuscript by moving the details of the analysis
of Hudson's algorithm and selective sweeps to appendices, and deleting
unnecessary text in several sections. We have added Table 1, which
summarises the features added to msprime since the 2016 paper.

@@

Software / API / innovations:
Clear separations of ancestry and mutation simulations
Improved methods for downstream analyses
Integration with other software including other simulators
Estimating run times of sim\_ancestry
ARG recording
Specific API for specifying demographic histories
Software development model

Implementations of external methods:
Selective sweeps
Instantaneous bottlenecks
Multiple merger coalescents
Discrete time Wright-Fisher
Gene conversion
Mutation models

As well as benchmarks against other software implementations.

The manuscript is well written and comprehensive - I do not find any major
deficiencies. This manuscript handles a wide range of topics with a high degree
of clarity. And I find the manuscript certainly reaches the criteria "... tools
of interest to a wide range of geneticists" and so is a good fit as an
investigation within the "Methods, Technology, \& Resources" subject.
\end{point}

\begin{reply}
Thank you for the comprehensive summary. We are delighted that these community
efforts are so well received.
\end{reply}

\begin{point}
However, I do think it would be possible to increase the clarity and usefulness
of the manuscript with some relatively small changes in organization.
\end{point}

\begin{reply}
Thank you for the detailed review and suggestions, which we hope have resulted
in an improved manuscript.
\end{reply}

@@

\begin{point}
As, is, the manuscript lacks a simple overview of the new capabilities of
version 1.0. Instead these are described in separate sections. The manuscript
might benefit from a short summary and or table of the software's basic
capabilities, API innovations, new functionality, etc. This may also provide a
way to structure the larger flow of the paper, which currently seems to touch
on the different topics in a somewhat random order.
\end{point}

\begin{reply}
We have added table 1 to summarise the new features.
\end{reply}

@@

\item Stated explicitly whether samples are haploid or diploid in the
figure captions, and updated axis labels also.

\item All figures now show the time averaged across replicates.

\end{itemize}

In addition figure 4 has been moved to the appendix, making the remaining
figures much more uniform in appearance and purpose (it is a special
case because it is showing how well theoretical predictions map
to observations, as well as those observations).
\end{reply}

@@

In addition to being highly efficient, the software engineering of msprime is
of the highest quality. The program includes extensive unit and validation
tests, and offers an API that simplifies worksflows and reduces the
inefficiency and sources of error associated with large intermediate text
files.

Kelleher has made every effort to engage with the community, with the result
that msprime now support many specialized features that have been contributed
by other developers. This has the important benefit of reducing the chances
that this software will lack support in future.

In short, this is an admirable project, and a shining example of successful
academic software development.
\end{point}

\begin{reply}
Thank you for the kind words, we are delighted that these community efforts are so well
received.
\end{reply}

@@

\begin{point}
The paper itself however is less polished. It is very long (37 pages / 1200
lines without appendix), mostly because it is in places overly detailed, and
seems to err on the side of completeness rather than conciseness or clarity.
\end{point}

\begin{reply}
Thank you for this well-founded criticism. We have reduced the length of the manuscript
by moving some material to appendices, cutting ``documentation-like'' text,
and tried to address specific points raised below. We hope that this has
resulted in an improved manuscript.
\end{reply}

@@

\begin{point}
For instance, lines 160-177 describe the data structure in some detail, while
this is published information. Line 192 remarks that "storage space is
dramatically reduced", and goes on to make this more precise and point out some
(fairly obvious) advantages in lines 193-202.
\end{point}

\begin{reply}
It is true that the details of the succinct tree sequence have been laid out
in earlier publications, and it is reassuring that the advantages are
obvious to you. However, for many of the intended audience these methods will
be new, and we feel it's important that this paper is a self-contained,
thorough, and convincing argument for the advantages of tree sequences and the
tskit library in simulation workflows.
Very many papers are still published using ms and other
simulators that use inefficient data formats, and it is these users that
we want to convince. We also hope to convince those that might be developing
their own simulators to take advantage of tskit and its extensive
functionality, rather then developing everything from scratch, as is the
classical and still dominant approach.
\end{reply}

@@

\begin{reply}
We have moved this detailed discussion to an appendix.
\end{reply}

\begin{point}
Another feature of the current paper is that a number of topics return several
times with minor variations. This is particularly true for algorithmic
efficiency, and the API, both of which are discussed multiple times. It is of
course true that algorithmic efficiency is the major distinguishing feature of
msprime; nevertheless, the focus of the current paper seems to be the rich
feature set, and the software development model, as well as the various APIs
that allow for a more integrated and less error prone workflow. I would suggest
to introduce separate sections for algorithmic efficiency and the API, and
discuss the various features (as you do now) in their own separate section, but
focus on these features from a user's rather than a developer's perspective.
\end{point}

\begin{reply}
The large number and diversity of features in msprime has made the
organisation of this
manuscript a major challenge, and the present layout of a series of
relatively self-contained topics arrived at after much experimentation.
Issues like interface design and algorithmic efficiency necessarily
cut across most of these topics. Separate sections on efficiency
and API design would need to refer to these other sections for context,
and would lead, we believe, to a longer and less cohesive document.

However, we do agree that some sections are too long and this criticism
has been very helpful to allow us see where content could be cut
without loss of information.
For example, we have cut out references to the API design in the
Simulating Mutations section, and substantially cut back the Demography
section (see next point).
\end{reply}

\begin{point}
Occasionally the paper reads more like documentation than a paper, such as when
the Demography class is introduced and several associated functions are
mentioned - this is not relevant for a paper in my opinion.
\end{point}
\begin{reply}
This is a very helpful observation, and helped us clarify where
text can be cut. We have substantially cut back the Demography section
and other areas where we were veering towards documentation.
\end{reply}

### 2021-11-24 response-to-reviewers.tex

\end{point}
\begin{reply}
Thank you for your suggestion, we have removed the number of lines of code
in the revised manuscript.
\end{reply}

\reviewersection

\begin{point}
This manuscript introduces a simulation tool, msprime version 1.0, and
describes features, additions, and performance of the simulator relative to the
earlier msprime software. I will begin by saying that I believe all researchers
in population genomics who use simulations have appreciated the quality work
that has been previously done in the msprime sphere. The contributions proposed
here clearly improve on a number of features, and that of course has value.
\end{point}
\begin{reply}
Thank you, we appreciate the kind words.
\end{reply}

\begin{point}
However, I did not find this work to be appropriate for Genetics, and would
rather view it as more appropriate for a journal like Bioinformatics. The
fundamental reason is that this basically appears to perform old
functionalities faster, and the added functionalities are modest. Again, both
are valuable, but this work doesn't address much larger biological /
statistical inference problems that are in need of development in this area,
which could justify publication in a journal like Genetics.
\end{point}

\begin{reply}
We respectfully disagree, and wonder if there is perhaps a simple
misunderstanding. While we absolutely agree that if this
manuscript were submitted as an ``Investigation'' to GENETICS,
it would need to make substantial methodological
advances of the type that you describe. However, we are submitting
the manuscript for consideration in the ``Methods, Technology, and Resources``
section, which has the following description:

\begin{quotation}
GENETICS welcomes manuscripts that describe genetic or analytic methods or
resources that are likely to have broad impact. They can be full length
research articles or Communications. The method or resource needs to be novel,
or be a significant advance in an existing method or resource. It should be of
considerable interest to a wide range of geneticists or of extraordinary
interest to a smaller group of geneticists. The method, resource or technology
should enable experiments that will allow investigators to address significant
biological questions and should be described well enough so others can
implement the method. The necessary reagents or resources need to be available
upon request.
\end{quotation}

We believe that this manuscript amply meets these requirements:

\begin{itemize}
\item The software is likely to have broad impact---the 2016 paper has
been cited 123 times so far in 2021 according to Google Scholar.
\item This is a significant advance over the 2016 paper, with a great deal of
new functionality (see the newly-added Table 1).
\item It is of interest to population geneticists who use simulations.
\item The functionality enables novel simulation types (e.g.,
$\Lambda$-coalescents) and greatly increases the scale over which
existing methods can be applied (e.g., bacterial simulations).
\end{itemize}

\end{reply}

@@

Torres et al. 2020. The temporal dynamics of background selection in
nonequilibrium populations. Genetics.

Johri et al. 2021. The impact of purifying and background selection on the
inference of population history: problems and prospects. MBE.

Murphy et al. 2021. Broad-scale variation in human genetic diversity levels is
predicted by purifying selection on coding and non-coding elements. bioRxiv.
\end{point}

\begin{reply}
We agree that neutral simulations are an approximation, and care needs to
be exercised in their use. We have added a sentence to introduction to
emphasise this point. However, as we have argued above, major methodological
advances of the type that you discuss (which we agree are extremely important!)
are beyond the scope of the current manuscript.
\end{reply}

\begin{point}
Relatedly, for the simulations of positive selection, while it has certainly
been common practice in the field for a few decades, it is rather odd to model
a single positively selected site on a background of all other neutral sites.
Positive selection may act on mutations in functional regions, and functional
regions will be largely shaped by purifying selection. Furthermore, the
Hill-Robertson effects between the modeled positively selected mutation and the
neglected array of negatively selected mutations is often non-trivial. In that
sense, the positive selection model is probably only of empirical relevance in
organisms with something approaching free recombination (to eliminate such
linkage effects), as in perhaps HIV or the like. But of course, these sorts of
organisms tend to have small, coding-dense genomes and experience very strong
selective pressures, thus returning to the first problem mentioned above that
regions free of background selection effects may not exist in the first place.
\end{point}
\begin{reply}
We agree that this is an important point, but again this is a deep
methodological issue that is beyond the scope of the current manuscript.
We have simply implemented methodology that has been used and
studied decades. Even if the goal is to demonstrate that such
approximations are deeply flawed, it is still necessary to be
able to simulate them efficiently and conveniently.
\end{reply}

@@

\begin{point}
In summary, I believe this to be a nice software addition that belongs as a
Note in a bioinformatics journal. The software solution that would represent an
important biological advance for Genetics would involve simulating more
realistic models as discussed above (as forward simulators are currently
capable of doing), but at a fast enough rate to be applicable to large-scale
simulation-based genomic inference.
\end{point}

\begin{reply}
Thank you for the detailed and interesting review---we agree that there
is much work to be done in the field of population genetic simulation!
As we have argued above, we believe the manuscript is entirely suitable
for publication in Genetics as a
``Methods, Technology, and Resources`` article.
\end{reply}
\end{document}

### 2021-11-24 paper.tex
Remove sentence about work being spread out

\section*{Acknowledgements}
We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
on the manuscript.
% alphabetised by author. Names spelled out if there's an initial collision
ADK was supported by NIH awards R01GM117241 and R01HG010774.
BE was supported by  DFG grant  273887127 through Priority Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
FB is funded by the Deutsche Forschungsgemeinschaft EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
Jerome Kelleher is supported by the Robertson Foundation.
SG acknowledges funding from the Canada Research Chairs Program, from the Canadian Institutes of Health Research PJT 173300, and from the Canadian Foundation for Innovation.

@@

What does this tell us about run time for simulating time-varying population sizes?
The argument above implies that the work is spread out relatively
evenly on the coalescent time scale.
Suppose that population size today is $N_1$,
while $T$ generations ago it was $N_2$.
Does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then run time will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the run time will depend critically on $T$---in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.

>>>>

\section*{Acknowledgements}
We acknowledge the contributions of Ivan Krukov who we consider
eligible for authorship, but were unable to contact for approval.
We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
on the manuscript.
% alphabetised by author. Names spelled out if there's an initial collision
ADK was supported by NIH awards R01GM117241 and R01HG010774.
BE was supported by  DFG grant  273887127 through Priority Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
FB is funded by the Deutsche Forschungsgemeinschaft EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
Jerome Kelleher is supported by the Robertson Foundation.
SG acknowledges funding from the Canada Research Chairs Program, from the Canadian Institutes of Health Research PJT 173300, and from the Canadian Foundation for Innovation.

@@

What does this tell us about run time for simulating time-varying population sizes?
Suppose that population size today is $N_1$,
while $T$ generations ago it was $N_2$.
Does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then run time will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the run time will depend critically on $T$---in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.
